Reasoning About a Machine with Local Capabilities

### 2018-05-05 toplas/paper.tex
Some modifications to the malloc section, explain that we use the LR machinery in the spec to allow malloc to depend on evolving internal state, describe some things a bit less defensively.

\begin{toplas}
  \section{Malloc}
  \label{sec:malloc}
In the examples we present in \sectionname~\ref{sec:examples}, we will assume
the existence of a trusted malloc routine so that it is possible to allocate new
memory. Our semantic model is not specific to a particular implementation of
malloc, so rather than providing the implementation, we provide a specification
that malloc must satisfy. This section can be skipped on first read through.

Malloc is assumed to be part of the trusted computing base, as mentioned in
\sectionname~\ref{sec:capab-mach-with}. A program with access to malloc should
be able to retain this access even if it invokes code it does not trust. In
order to achieve this, a capability for malloc must be global. It is also
important that malloc remains trustworthy to all entities on the machine
throughout execution. That is, malloc should encapsulate its local state and
not depend on any external code. To make sure that malloc cannot be tampered
with directly through the malloc capability, we require it to have enter
permission.

The requirements so far are syntactic and prevent direct tampering in the sense
that the capability intended for invoking malloc only can be used for
invocation. This does, however, not prevent the implementation of malloc from
colluding with other code. Specifically, malloc could depend on the state of a
piece of memory not just under the control of malloc. We also need to make sure
that malloc actually behaves as intended, that is, when malloc is invoked with a
length argument, then it actually returns a capability for a fresh piece of
memory of that size. It should be fresh in the sense that malloc has not given
out a capability for any part of that memory before and will not do it in the
future. Finally, we also need to make sure that malloc behaves correctly when
given a nonsensical length argument such as a negative integer or a capability.
These requirements relate to the behaviour of malloc, so we cannot hope to
express them syntactically. Instead we will use the semantic model defined in
Section~\ref{sec:logical-relation} to express this part of the specification.

For the semantic requirements on malloc, the specification will
require the existence of some initial region for malloc. Malloc does not
disappear from the system, so the malloc region must be permanent. This also
means that the malloc specification should guarantee that use of malloc is safe
at any point of execution no matter who has invoked it. To this end, the
requirement on the initial malloc region should also hold true in any private
future world.
% priv future means that malloc should not depend on local capabilities?

The implementation of malloc should not depend on any memory outside the
implementation of malloc. This means that malloc should not have any
capabilities in its internals that govern other programs. As explained in
\sectionname~\ref{subsec:worlds}, a region's state interpretation function
determines what memory segments it permits with respect to some world. The state
interpretation function of malloc should permit the same regions with respect to
a world with the malloc region as it would with respect to a world that only
contains the malloc region. In other words, the remaining regions in the world
should not influence which regions malloc accepts. Malloc should retain this
property throughout execution, so it must hold true for any private future
region of the initial malloc region.

The malloc specification will also dictate what malloc should do when invoked.
Malloc can be invoked with either a valid or an invalid length argument. In the
case where malloc is invoked with an invalid argument, that is, a negative
integer or a capability, we simply require malloc to fail. This part of the
specification does not rely on the malloc region, as failing is considered
safe behaviour. When malloc is invoked with a valid length, then malloc should
return a fresh memory segment of the correct length.
Malloc may have been invoked a number of times before any given invocation, so
the internal data structures of malloc may be in a different state than they
were initially.
We know that the memory of malloc is in a state consistent with the
protocol dictated by the initial malloc region in the sense that we can assume
that malloc is invoked in a configuration where part of it satisfies a private
future region of the initial malloc region. Knowing the state of the malloc
region allows us to express what we mean by a \emph{fresh} region. This means
that malloc should not keep any capabilities for the freshly allocated region.
To ensure this, we require the footprint of malloc after the call to be disjoint
from the freshly allocated memory. Further, the footprint should satisfy a
public future region of the malloc region the invocation happened in. This means
that it cannot depend on the freshly allocated region. It further allows
subsequent calls to malloc as it provides one of the requirements in the malloc
specification.

Malloc is part of the trusted computing base. If we cannot trust malloc, then we
cannot use the memory it allocates as we have no idea whether it is aliased by
some untrusted program. Because of this assumption, we also require malloc to
promise that the non-argument registers of the register-file remain unchanged
after the call. This allows the caller to keep private capabilities in the
register file (strictly speaking this is not necessary as we can use our calling
convention to protect our local state).

\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff
  \[  
    \begin{array}{l}
      c_\malloc = ((\entry,\glob),\_,\_,\_) \land\\
      \exists \iota_{\malloc,0} \ldotp \\
      \quad \iota_{\malloc,0}.v = \perma \; \land \\
      \quad \left(
        \begin{array}{l}
          \forall \iota' \futurestr \iota_{\malloc,0}, W,i \ldotp\\
          \quad W(i)=\iota' \Rightarrow \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
        \end{array}
        \right) \wedge \\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp \forall \ms_{\var{footprint}}, \hsframe \in \HeapSegments \ldotp \\
        \quad \forall i, n, \size \in \nats \ldotp \forall
        w_{\var{ret}} \in \Words \ldotp \\
        \quad \quad \forall \iota_\malloc \futurestr \iota_{\malloc,0} \land \\
        \quad \quad \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
        \qquad \memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
        \qquad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
        \qquad \Rightarrow \\
        \qquad \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
        \qquad \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
        \qquad \qquad \quad \Phi \step[j] \Phi' \land \\
        \qquad \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
        \qquad \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
        \qquad \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
        \qquad \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
        \qquad \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
        \qquad \qquad \quad \size - 1 = e'-b'
      \end{array}
      \right) \land\\
      \quad \left(
      \begin{array}{l}
        \forall \Phi \in \ExecConfs \ldotp\\
        \quad(\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
        \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
        \qquad \Rightarrow \exists j \in \nats \ldotp \Phi \step[j] \failed
      \end{array}
        \right)
    \end{array}
  \]
\end{definition}
As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously can not
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

Malloc should not just be available to trusted programs, but also to possibly
malicious programs. The specification should therefore entail that the malloc
capability is safe which would make it safe to let even malicious programs have
access to it.
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For all $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

The focus of this work is not to give a definitive specification and
implementation of malloc which is why we do give a malloc implementation. We
are, however, confident that it is possible to make an implementation of malloc
that satisfies the malloc specification in Definition~\ref{spec:malloc}. There
are in particular two simplifications in our system that makes things easy:
First, we do not consider deallocation of memory which means that the data
structure malloc uses to keep track of free memory does not have to handle
reclaimed memory. Second, the malloc specification does nor permit malloc to run
out of memory and thus prevent allocation. This is possible on our simple
capability machine because it has an infinite address space. The capability with
an infinite range of authority would of course need to be part of malloc, but it
could also double as the data structure that keeps track of free memory.
\itoplassug{If we have references for malloc being difficult, we could add that
  to the text.}
%\itoplassug{Add the assumption that the malloc capability in examples
%  satisfies specification.}

>>>>

\begin{toplas}
  \section{Malloc}
  \label{sec:malloc}
  In the examples we present in \sectionname~\ref{sec:examples}, we will assume
  the existence of a trusted malloc routine so that both the trusted code and
  the adversary can allocate new memory. Malloc is considered part of the
  trusted computing base, as mentioned in
  \sectionname~\ref{sec:capab-mach-with}. This is unavoidable: if we cannot
  trust malloc, then we cannot use the memory it allocates as we have no idea
  whether it is aliased by some untrusted program.

  Our semantic model is not specific to a particular implementation of malloc,
  so rather than providing the implementation, we provide a specification that
  malloc must satisfy. The specification is a bit technical, but really just
  expresses standard expectations of how malloc should behave and this section
  can safely be skipped on first read.

We require a global capability for
invoking malloc (because if the capability were local, then a program with
access to malloc would have to give up this access when invoking untrusted
code). The capability is assumed to have enter permission, so that malloc can
protect its internal state even when the capability is shared with untrusted
code.

In addition to these syntactic requirements, we also specify standard intuitive
expectations for how malloc behaves. When malloc is invoked with a length
argument, then it will return a capability for a fresh piece of memory of that
size. It should be fresh in the sense that malloc has not already given out a
capability for any part of that memory before and will not do so in the future.
Finally, we also need to make sure that malloc behaves correctly when given a
nonsensical length argument such as a negative integer or a capability. However,
formulating these requirements is harder than one might expect. The problem is
that we want to allow malloc implementations to rely on internal state and
invariants on that state and it is reasonable to expect both this internal state
and the invariants to change over time. To express that malloc should behave
according to its specifications, but only if its internal state satisfies its
current invariants in an executing system, we will use the semantic model
defined in Section~\ref{sec:logical-relation} to express our specification.

To allow malloc implementations to rely on internal state and invariants for
that state, we assume an initial region for malloc. The region is assumed to be
permanent (since safety of the global malloc capability will depend on its
presence).
% DOMI: I don't understand what the following uncommented text means
% This also means that the malloc specification should guarantee that
% use of malloc is safe at any point of execution no matter who has invoked it. To
% this end, the requirement on the initial malloc region should also hold true in
% any private future world.
% priv future means that malloc should not depend on local capabilities?
Furthermore, we want to express that malloc does not depend on any other memory
than its own internal state. This is expressed by a restriction on the malloc
region's state interpretation function, which (as explained in
\sectionname~\ref{subsec:worlds}) defines what memory segments it permits in a
given world. We require that the accepted memory segments only depend on the
presence of the malloc region itself, i.e.\ that in any world, the same memory
segments are accepted if we remove all regions except the malloc region. This
property should continue to hold throughout execution, so it must hold true for
any private future region of the initial malloc region.

The malloc specification then dictates what malloc should do when invoked in a
memory with its internal state valid according to the malloc region (some future
evolution of the initial malloc region). Malloc is invoked with a length
argument. If this argument is invalid, that is, a negative integer or a
capability, we simply require malloc to fail. This part of the specification
does not actually rely on the malloc region: for simplicity, we assume malloc
does not need its internal state to check the argument. When malloc is invoked
with a valid length, then it should return a fresh memory segment of the correct
length. This segment is required to come from the footprint of malloc, i.e.\ the
memory owned by the malloc region before the call. After malloc returns, we
require the malloc region to have evolved (according to the public future world
relation) to a new state where the new footprint is disjoint from the allocated
memory. This implies that future invocations of malloc can never return
previously-allocated memory.

For convenience, we also require that malloc returns the non-argument registers
of the register-file unchanged after the call. This allows the caller to keep
private capabilities in the register file, without having to protect them by
storing them in a private stack frame.

\begin{definition}[Malloc Specification]
  \label{spec:malloc}
  $c_\malloc$ satisfies the specification for malloc iff the following conditions hold:
  \begin{itemize}
  \item $c_\malloc = ((\entry,\glob),\_,\_,\_)$
  \item There exists a $\iota_{\malloc,0}$ such that
    \begin{itemize}
    \item $\iota_{\malloc,0}.v = \perma$
    \item For all $\iota' \futurestr \iota_{\malloc,0}, W,i$ with $W(i)=\iota'$, we have that
      \begin{equation*}
        \iota'.H (\iota'.s) (\xi^{-1}(W)) = \iota'.H (\iota'.s) (\xi^{-1}([i \mapsto W(i)]))
      \end{equation*}
    \item  
        For all $\Phi \in \ExecConfs$, $\ms_{\var{footprint}}, \hsframe \in \HeapSegments$, $i, n, \size \in \nats$, $w_{\var{ret}} \in \Words$, $\iota_\malloc \futurestr \iota_{\malloc,0}$, we have that
      \[
      \begin{array}{l}
        \text{If }
        \memheap = \ms_{\var{footprint}} \uplus \hsframe \land \heapSat[\ms_{\var{footprint}}]{n}{[i \mapsto \iota_\malloc]} \land \\
        \quad\memreg(r_1) = \size \land \size \geq 0 \land  \memreg(r_0) = w_{\var{ret}} \land \\
        \quad \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\ %and this needs to change with ABI.
        \text{Then, } \\
        \quad \exists \Phi' \in \ExecConfs \ldotp \exists \ms_{\var{footprint}}', \ms_{\var{alloc}} \in \HeapSegments\ldotp\\
        \qquad \exists j \in \nats \ldotp j > 0 \land \exists b',e'\in \Addrs \ldotp \exists \iota_\malloc' \in \Regions \ldotp \\
        \qquad \quad \Phi \step[j] \Phi' \land \\
        \qquad \quad \memheap[\Phi']=\ms_{\var{footprint}}' \uplus \hs_{\var{alloc}} \uplus \hsframe \land\\
        \qquad \quad \iota_{\malloc}' \futurewk \iota_\malloc \land \\
        \qquad \quad \heapSat[\ms_{\var{footprint}}']{n-j}{[i \mapsto \iota_\malloc']} \land \\
        \qquad \quad \dom(\hs_{\var{alloc}}) = [b',e'] \land \forall a \in [b',e']\ldotp \hs_{\var{alloc}}(a) = 0  \land \\
        \qquad \quad \memreg[\Phi'] = \memreg[\Phi]\update{\pcreg}{\updatePcPerm{w_{\var{ret}}}}\update{r_1}{((\rwx,\glob),b',e',b')} \land \\
        \qquad \quad \size - 1 = e'-b'
      \end{array}
      \]
    \item
      For all $\Phi \in \ExecConfs$,
      \[  
          \begin{array}{l}
            \text{If } (\memreg(r_1) \notin \ints \vee \memreg(r_1) < 0) \land
            \memreg(\pcreg) = \updatePcPerm{c_\malloc} \\
            \text{Then } \exists j \in \nats \ldotp \Phi \step[j] \failed
          \end{array}
      \]
\end{itemize}
\end{itemize}
\end{definition}
As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously can not
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

Malloc should not just be available to trusted programs, but also to possibly
malicious programs. This is safe as it follows from the specification that the malloc
capability is always safe in a world with the malloc region:
\begin{lemma}[Malloc is safe to pass to adversary]
  \label{lem:malloc-in-vr}
  For all $c_\malloc$ that satisfies the specification for malloc with region $\iota_{\malloc,0}$,  if $W(r) \futurestr \iota_{\malloc,0}$, then
  $\npair{c_\malloc} \in \stdvr(W)$ for all $n$.
\end{lemma}

The focus of this work is not to give a definitive specification and
implementation of malloc which is why we do not give a malloc implementation. We
are, however, confident that it is possible to make an implementation of malloc
that satisfies the malloc specification in Definition~\ref{spec:malloc}. There
are in particular two simplifications in our system that makes things easy:
First, we do not consider deallocation of memory which means that the data
structure malloc uses to keep track of free memory does not have to handle
reclaimed memory. Second, the malloc specification does not permit malloc to run
out of memory and thus refuse allocation. This is possible on our simple
capability machine because it has an infinite address space. An initial
capability with an infinite range of authority would of course need to be part
of malloc, but it could also double as the data structure that keeps track of
free memory. \itoplassug{If we have references for malloc being difficult, we
  could add that to the text.}
%\itoplassug{Add the assumption that the malloc capability in examples
%  satisfies specification.}
